---
title: "chapter3simu"
format: pdf
editor: visual
---

## Chapter 3 Simulations

Focus on Simulations and Expectations of Poisson Processes

## 3.2 Simulations in R

### Necessary Packages:

```{r}
library(diagram)  
suppressWarnings(library(expm))   
library(markovchain)
```

### Simulation 3.1: Exponential Interarrivals

Using inverse of the cdf of standard uniform variables to get exp random variables with the correct mean.

```{r}
 #specifying parameters
 lambda<- 2
 njumps<- 20
 #defining states
 N<- 0:njumps
 #setting timeas vector
 time<- c()
 #setting initialvalue fortime
 time[1]<- 0
 #specifying seed
 set.seed(333422)
 #simulating trajectory
 for (i in 2:(njumps+1))
 time[i]<- time[i-1]+round((-1/lambda)*log(runif(1)),2)
 #plotting trajectory
 #type="n"draws emptyframewithnograph
 plot(time, N,type="n", xlab="Time",ylab="State", panel.first
 =grid())
 segments(time[-length(time)], N[-length(time)],time[-1]-0.07,
 N[-length(time)], lwd=2,col="blue")
 points(time, N,pch=20, col="blue")
 points(time[-1], N[-length(time)],pch=1, col="blue")
```

### Simulation 3.2: Uniform Order Statistics

Simulating waiting times:

1\) Fix t and generate N(t) -\> Pois( $\lambda$ t)

2\) Generate N(t) standard uniform random variables $U_1 , ... , U_{N(t)}$

3\) Order $U_1 , ... , U_{N(t)}$ in increasing order to get ordered set $O_{N(t)}$

4\) Multiply ordered set by t to obtain set of event times $S_1 = t*O_{N(t)}$

5\) Define states of Poisson Process as: N(0) = 0, $N(S_1) = 1$, ... $N(S_{N(t)}) = N(t)$

6\) Plot states against time

```{r}
#specifying parameters
 t<- 10
 lambda<- 2
 #specifying seed
 set.seed(32114)
 #generating N(t)
 njumps<- rpois(1,lambda*t)
 #defining states
 N<- 0:njumps
 #generating N(t)standard uniforms
 u<- c()
 u[1]<- 0
 for(i in 2:(njumps+1))
 u[i]<- runif(1)
 #computing eventtimes
 time<- t*sort(u)
 #plotting trajectory
 plot(time, N,type="n", xlab="Time",ylab="State", panel.first
 =grid())
 segments(time[-length(time)], N[-length(time)],time[-1]-0.07,
 N[-length(time)], lwd=2,col="blue")
 points(time, N,pch=20, col="blue")
 points(time[-1], N[-length(time)],pch=1, col="blue")
```

## 3.3 Applications of Poisson Process

### Application 3.1: Seismology

Occurrences of earthquakes can be modeled by poisson processes:

(missing the file)

```{r}
eq.data<- read.csv(file="./earthquakedata2012-2018.csv",
 header=TRUE, sep=",")
 #creating date-time variable
 datetime<- as.POSIXct(paste(as.Date(eq.data$DATE), eq.data$TIME))
 #computing lag
 datetime.lag<- c(0,head(datetime,-1))
 #computing interarrival times (in hours)
 int.time<- (as.numeric(datetime)-as.numeric(datetime.lag))/3600
 #removing first value
 int.time<- int.time[-1]
 #removing immediate aftershocks (within 3 hours)
 int<- int.time[int.time>3]
 #plotting histogram
 hist(int, main="", col="dark magenta", xlab="Interarrival Time")
```

Bins of Different Interarrival Times, exp(), chi-square check:

```{r}
#binning interarrivaltimes
 binned.int<- as.factor(ifelse(int<40,"1",
 ifelse(int>=40& int<80,"2",ifelse(int>=80 &int<120,"3",
 ifelse(int>=120& int<160,"4",ifelse(int>=160 &int<200,"5",
 ifelse(int>=200& int<240,"6","7")))))))
 #computing observedfrequencies
 obs<- table(binned.int)
 #estimating meanfor exponentialdistribution
 mean.est<- mean(int)
 #computing expectedfrequencies
 exp<- c(1:7)
 exp[1]<- length(int)*(1-exp(-40/mean.est))
 exp[2]<- length(int)*(exp(-40/mean.est)-exp(-80/mean.est))
 exp[3]<- length(int)*(exp(-80/mean.est)-exp(-120/mean.est))
 exp[4]<- length(int)*(exp(-120/mean.est)-exp(-160/mean.est))
 exp[5]<- length(int)*(exp(-160/mean.est)-exp(-200/mean.est))
 exp[6]<- length(int)*(exp(-200/mean.est)-exp(-240/mean.est))
 exp[7]<- length(int)*exp(-240/mean.est)
 obs
round(exp,1)
#computing chi-squaredstatistic
print(chi.sq<- sum((obs-exp)^2/exp))
#computing p-value
print(p.value<- 1-pchisq(chi.sq,df=5))
```

P-value \> 0.05 so earthquake follows poisson process

## Exercises

### Exercise 3.6

A\) Average \# of distressed road on 10 mile stretch

```{r}
10 / 2.8
```

B\) Simulate 30 distressed surface areas

```{r}
#specifying parameters
 lambda<- 2.8
 njumps<- 30
 #defining states
 N<- 0:njumps
 #setting timeas vector
 time<- c()
 #setting initialvalue fortime
 time[1]<- 0
 #specifying seed
 set.seed(333422)
 #simulating trajectory
 for (i in 2:(njumps+1))
 time[i]<- time[i-1]+round((-1/lambda)*log(runif(1)),2)
 #plotting trajectory
 #type="n"draws emptyframewithnograph
 plot(time, N,type="n", xlab="Time",ylab="State", panel.first
 =grid())
 segments(time[-length(time)], N[-length(time)],time[-1]-0.07,
 N[-length(time)], lwd=2,col="blue")
 points(time, N,pch=20, col="blue")
 points(time[-1], N[-length(time)],pch=1, col="blue")
 
 time[30]
```

15.4 miles

c\) Finding 30 areas in 10 miles

```{r}
#specifying parameters
 lambda<- 2.8
 njumps<- 30
 #defining states
 N<- 0:njumps
 #setting timeas vector
 time<- c()
 #setting initialvalue fortime
 time[1]<- 0
 #specifying seed
 set.seed(333422)
 #simulating trajectory
 limit <- 10
 time[30] <- limit+1
 while (time[30] > limit) {
 time[1] <- 0
 for (i in 2:(njumps+1))
 time[i]<- time[i-1]+round((-1/lambda)*log(runif(1)),2)
 }
 #plotting trajectory
 #type="n"draws emptyframewithnograph
 plot(time, N,type="n", xlab="Time",ylab="State", panel.first
 =grid())
 segments(time[-length(time)], N[-length(time)],time[-1]-0.07,
 N[-length(time)], lwd=2,col="blue")
 points(time, N,pch=20, col="blue")
 points(time[-1], N[-length(time)],pch=1, col="blue")
 
 time[30]
```
